Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-23 paper/main.tex
edits up through pg 14

A common approach to assessing a student's knowledge is to present them with a set of quiz
questions, calculate the proportion they answer correctly, and provide
them with feedback in the form of a simple numeric or letter grade. While such a
grade can provide \textit{some} indication of whether the student has
mastered the to-be-learned material, any univariate measure of performance on a
complex task sacrifices certain relevant information, risks conflating underlying
factors, and so on. For example, consider the relative utility of the
imaginary map described above that characterizes a student's knowledge
in detail, versus a single annotation saying that the student answered 85\% of
their quiz questions correctly, or that they received a `B'. Here, we show that
the same quiz data required to compute proportion-correct scores or letter
grades can instead be used to obtain far more detailed insights into what a
student knew at the time they took the quiz.

@@

Our primary research goal is to advance our understanding of what it means to
acquire deep, real-world conceptual knowledge. Traditional laboratory
approaches to studying learning and memory (e.g., list-learning studies) often
draw little distinction between memorization and understanding. Instead, these
studies typically focus on whether information is effectively encoded or
retrieved, rather than whether the information is \textit{understood}.
Approaches to studying conceptual learning, such as category learning
experiments, can begin to investigate the distinction between memorization and
understanding, often by training participants to distinguish arbitrary or
random features in otherwise meaningless categorized stimuli. However the
objective of real-world training, or learning from life experiences more
generally, is often to develop new knowledge that may be applied in
\textit{useful} ways in the future. In this sense, the gap between modern
learning theories and modern pedagogical approaches that inform classroom
learning strategies is enormous: most of our theories about \textit{how} people
learn are inspired by experimental paradigms and models that have only
peripheral relevance to the kinds of learning that students and teachers
actually seek~\citep{Macl05, HallGree08}. To help bridge this gap, our study
uses course materials from real online courses to inform, fit, and test models
of real-world conceptual learning. We also provide a demonstration of how our
models can be used to construct ``maps'' of what students know, and how their
knowledge changes with training. In addition to helping to visually capture knowledge
(and changes in knowledge), we hope that such maps might lead to real-world
tools for improving how we educate. Taken together, our work shows that
existing course materials and evaluative tools like short multiple-choice
quizzes may be leveraged to gain highly detailed insights into what students
know and how they learn.

@@

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on our participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted the two lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same (per
instructor annotations) Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overlap
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (Lectures~1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics not specifically presented in either
video (see Supplementary Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(Lecture~1, Lecture~2, and general physics knowledge) on each of the three
quizzes. Quiz~1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz~2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., Lecture~1), and Quiz~3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., Lecture~2).

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents, and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time span. We treated the set
of text snippets (across all of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix that
described the unique mixture of broad themes from both lectures reflected in
each window's text. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or question set) reflects about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the Lecture~1 video
and questions ($r(13) = 0.824,~p<0.001$, 95\% CI~$= [0.696,~0.973]$) and the Lecture~2
video and questions ($r(13) = 0.801,~p<0.001$, 95\% CI~$= [0.539,~0.958]$). 
However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., Lecture~1 video vs.~Lecture~2 questions; Lecture~2 video
vs.~Lecture~1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figure~\ref{fig:topics} and Supplementary Figure~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondences (at a
coarse scale) between the lectures and questions.

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether the text embeddings could be used to
automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors from each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also qualitatively examined the
best-matching intervals for each question by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question probed and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

\begin{figure}[tp] \centering
\includegraphics[width=\textwidth]{figs/lecture-question-similarity}

@@

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail computing the
proportion of correctly answered questions. But if two students receive
identical scores on an exam, might our modeling framework help us to gain more
nuanced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to fill in the ``gaps'' in the two students'
understandings, we might do well to focus on concept $A$ for the first student,
but to also add in materials pertaining to concepts $B$ and $C$ for the second
student. In other words, raw ``proportion-correct'' measures may capture
\textit{how much} a student knows, but not \textit{what} they know. We wondered
whether our modeling framework might enable us to (formally and automatically)
infer participants' knowledge at the scale of individual concepts (e.g., as
captured by a single moment of a lecture).

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \nameref{subsec:traces}). Essentially, the
estimated knowledge at the coordinate is given by the weighted average
proportion of quiz questions the participant answered correctly, where the
weights reflect how much each question is ``about'' the content at $x$. When we
apply this approach to estimate the participant's knowledge about the content
presented in each moment of each lecture, we can obtain a detailed timecourse
describing how much ``knowledge'' the participant has about any part of the
lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also apply
this approach separately for the questions from each quiz the participants took
throughout the experiment. From just a few questions per quiz (see Sec.~\ref{subsec:traces}), we obtain a
high-resolution snapshot (at the time each quiz was taken) of what the
participants knew about any moment's content, from either of the two lectures
they watched (comprising a total of 1,100 samples across the two lectures).

@@

    \caption{\textbf{Estimating moment-by-moment knowledge acquisition.}
    \textbf{A. Moment-by-moment knowledge about the \textit{Four Fundamental
    Forces}}. Each trace displays the weighted proportion of correctly answered
    questions about the content reflected in each moment of the lecture (see
    \nameref{subsec:traces}), using responses from one
    quiz (color). The traces are averaged across participants. \textbf{B.
    Average estimated knowledge about the \textbf{Four Fundamental Forces}.}
    Each bar displays the across-timepoint average knowledge, estimated using
    the responses to one quiz's questions. \textbf{C. Moment-by-moment
    knowledge about the \textit{Birth of Stars}}. The panel is in the same
    format as Panel A, but here the knowledge estimates are for the
    moment-by-moment content of the \textit{Birth of Stars} lecture. \textbf{D.
    Average estimated knowledge about the \textit{Birth of Stars}.} The panel
    is in the same format as Panel B, but here the knowledge estimates are for
    the content of the \textit{Birth of Stars} lecture. All panels: error
    ribbons and error bars denote 95\% confidence intervals, estimated across
    participants.}

@@

Of course, even though the timecourses in
Figure~\ref{fig:knowledge-timeseries}A and C provide detailed
\textit{estimates} about participants' knowlege, those estimates are only
\textit{useful} to the extent that they accurately reflect what participants
actually know. As one sanity check, we anticipated that the knowledge estimates
should show a content-specific ``boost'' in participants' knowledge after
watching each lecture. In other words, if participants learn about each
lecture's content when they watch each lecture, the knowledge estimates should
reflect that. After watching the \textit{Four Fundamental Forces} lecture,
participants should show more knowledge for the content of that lecture than
they had before, and that knowledge should persist for the remainder of the
experiment. Specifically, knowledge about that lecture's content should be
relatively low when estimated using Quiz~1 responses, but should increase when
estimated using Quiz~2 or 3 responses (Fig.~\ref{fig:knowledge-timeseries}B).
Indeed, we found that participants' estimated knowledge about the content of
the \textit{Four Fundamental Forces} was substantially higher on Quiz~2 versus
Quiz~1 ($t(49) = 8.764,~p < 0.001$) and on Quiz~3 versus Quiz~1 ($t(49) =
10.519,~p < 0.001$). We found no reliable differences in estimated knowledge
about that lecture's content on Quiz~2 versus 3 ($t(49) = 0.160,~p = 0.874$).
Similarly, we hypothesized (and subsequently confirmed) that participants
should show more estimated knowledge about the content of the \textit{Birth of
Stars} lecture after (versus before) watching it
(Fig.~\ref{fig:knowledge-timeseries}D). Specifically, since participants
watched that lecture after taking Quiz~2 (but before Quiz~3), we hypothesized
that their knowledge estimates should be relatively low on Quizzes~1 and 2, but
should show a ``boost'' on Quiz~3. Consistent with this prediction, we found no
reliable differences in estimated knowledge about the \textit{Birth of Stars}
lecture content on Quizzes~1 versus 2 ($t(49) = 1.013,~p = 0.316$), but the
estimated knowledge was substantially higher on Quiz~3 versus 2 ($t(49) =
10.561,~p < 0.001$) and Quiz~3 versus 1 ($t(49) = 8.969,~p < 0.001$).

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should carry some
predictive information about whether the participant is likely to answer the
question correctly or incorrectly. We developed a statistical approach to test this claim. 
For each question in turn, for each
participant, we used Equation~\ref{eqn:prop} to estimate (using all
\textit{other} questions from the same quiz, from the same participant) the
participant's knowledge at the held-out question's embedding coordinate. For
each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of each \textit{correctly} answered
question, and another for the estimated knowledge at the coordinates of each
\textit{incorrectly} answered question (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be applied to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with $k = 100$ topics. We hoped that increasing the number of
topics from 15 to 100 might help us to generalize the knowledge predictions.
(Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.)
As in our other analyses, we resampled each lecture's topic trajectory to 1~Hz
and also projected each question into a shared text embedding space.

>>>>

A common approach to assessing a student's knowledge is to present them with a set of quiz
questions, calculate the proportion they answer correctly, and provide
them with feedback in the form of a simple numeric or letter grade. While such a
grade can provide \textit{some} indication of whether the student has
mastered the to-be-learned material, any univariate measure of performance on a
complex task sacrifices certain relevant information, risks conflating underlying
factors, and so on. For example, consider the relative utility of the
theoretical map described above that characterizes a student's knowledge
in detail, versus a single annotation saying that the student answered 85\% of
their quiz questions correctly, or that they received a `B'. Here, we show that
the same quiz data required to compute proportion-correct scores or letter
grades can instead be used to obtain far more detailed insights into what a
student knew at the time they took the quiz.

@@

Our primary research goal is to advance our understanding of what it means to
acquire deep, real-world conceptual knowledge. Traditional laboratory
approaches to studying learning and memory (e.g., list-learning studies) often
draw little distinction between memorization and understanding. Instead, these
studies typically focus on whether information is effectively encoded or
retrieved, rather than whether the information is \textit{understood}.
Approaches to studying conceptual learning, such as category learning
experiments, can begin to investigate the distinction between memorization and
understanding, often by training participants to distinguish arbitrary or
random features in otherwise meaningless categorized stimuli~\citep{ReilEtal82, Este86a, Este86b, GlucEtal02, AshbMadd05, HulbNorm15}. However the
objective of real-world training, or learning from life experiences more
generally, is often to develop new knowledge that may be applied in
\textit{useful} ways in the future. In this sense, the gap between modern
learning theories and modern pedagogical approaches that inform classroom
learning strategies is enormous: most of our theories about \textit{how} people
learn are inspired by experimental paradigms and models that have only
peripheral relevance to the kinds of learning that students and teachers
actually seek~\citep{Macl05, HallGree08}. To help bridge this gap, our study
uses course materials from real online courses to inform, fit, and test models
of real-world conceptual learning. We also provide a demonstration of how our
models can be used to construct ``maps'' of what students know, and how their
knowledge changes with training. In addition to helping to visually capture knowledge
(and changes in knowledge), we hope that such maps might lead to real-world
tools for improving how we educate. Taken together, our work shows that
existing course materials and evaluative tools like short multiple-choice
quizzes may be leveraged to gain highly detailed insights into what students
know and how they learn.

@@

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted the two lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same
Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overlap
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (Lectures~1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics concepts not specifically presented in either
video (see Supplementary Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(Lecture~1, Lecture~2, and general physics knowledge) on each of the three
quizzes. Quiz~1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz~2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., Lecture~1), and Quiz~3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., Lecture~2).

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents, and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time span. We treated the set
of text snippets (across all of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix
describing the unique mixture of broad themes from both lectures reflected in
each window's text. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or question set) reflects about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the Lecture~1 video
and questions ($r(13) = 0.824,~p<0.001$, 95\% CI~$= [0.696,~0.973]$) and the Lecture~2
video and questions ($r(13) = 0.801,~p<0.001$, 95\% CI~$= [0.539,~0.958]$). 
Simultaneously, as reported in Figure~\ref{fig:topics}B, the variability in
topic expression across \textit{different} videos and lecture-specific
questions (i.e., Lecture~1 video vs.~Lecture~2 questions; Lecture~2 video
vs.~Lecture~1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figure~\ref{fig:topics} and Supplementary Figure~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondences (at a
coarse scale) between the lectures and questions.

While an individual lecture may be organized around a single broad theme at a
coarse scale, at a finer scale, each moment of a lecture typically covers a
narrower range of content. Given the correspondence we found between the variability
in topic expression across moments of each lecture and questions from its
corresponding set (Fig.~\ref{fig:topics}), we wondered whether the text embedding model
might additionally capture these conceptual relationships at a finer scale.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether the text embeddings could be used to
automatically identify the ``matching'' moment(s) in the lecture. To explore this,
we computed the correlation between each question's topic weights and the topic weights
for each second of its corresponding lecture, and found that each question appeared to be
temporally specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also qualitatively examined the
best-matching intervals for each question by comparing the question's text
to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question probed and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth]{figs/lecture-question-similarity}

@@

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail computing the
proportion of correctly answered questions. But if two students receive
identical scores on an exam, might our modeling framework help us to gain more
nuanced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to help these two students fill in the ``gaps'' in their
understandings, we might do well to focus specifically on concept $A$ for the first student,
but to also add in materials pertaining to concepts $B$ and $C$ for the second
student. In other words, raw ``proportion-correct'' measures may capture
\textit{how much} a student knows, but not \textit{what} they know. We wondered
whether our modeling framework might enable us to (formally and automatically)
infer participants' knowledge at the scale of individual concepts (e.g., as
captured by a single moment of a lecture).

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \nameref{subsec:traces}). Essentially, the
estimated knowledge at coordinate $x$ is given by the weighted average
proportion of quiz questions the participant answered correctly, where the
weights reflect how much each question is ``about'' the content at $x$. When we
apply this approach to estimate the participant's knowledge about the content
presented in each moment of each lecture, we can obtain a detailed timecourse
describing how much ``knowledge'' the participant has about any part of the
lecture. As shown in Figure~\ref{fig:knowledge-timeseries}A and C, we can apply
this approach separately for the questions from each quiz participants took
throughout the experiment. From just a few questions per quiz (see~\nameref{subsec:traces}), we obtain a
high-resolution snapshot (at the time each quiz was taken) of what the
participants knew about any moment's content, from either of the two lectures
they watched (comprising a total of 1,100 samples across the two lectures).

@@

    \caption{\textbf{Estimating moment-by-moment knowledge acquisition.}
    \textbf{A. Moment-by-moment knowledge about the \textit{Four Fundamental
    Forces}}. Each trace displays the weighted proportion of correctly answered
    questions about the content reflected in each moment of the lecture (see
    \nameref{subsec:traces}), using responses from a single
    quiz (color). The traces are averaged across participants. \textbf{B.
    Average estimated knowledge about the \textbf{Four Fundamental Forces}.}
    Each bar displays the across-timepoint average knowledge, estimated using
    the responses to one quiz's questions. \textbf{C. Moment-by-moment
    knowledge about the \textit{Birth of Stars}}. The panel is in the same
    format as Panel A, but here the knowledge estimates are for the
    moment-by-moment content of the \textit{Birth of Stars} lecture. \textbf{D.
    Average estimated knowledge about the \textit{Birth of Stars}.} The panel
    is in the same format as Panel B, but here the knowledge estimates are for
    the content of the \textit{Birth of Stars} lecture. All panels: error
    ribbons and error bars denote 95\% confidence intervals, estimated across
    participants.}

@@

While the timecourses in
Figure~\ref{fig:knowledge-timeseries}A and C provide detailed
\textit{estimates} about participants' knowlege, these estimates are of course only
\textit{useful} to the extent that they accurately reflect what participants
actually know. As one sanity check, we anticipated that the knowledge estimates
should reflect a content-specific ``boost'' in participants' knowledge after
watching each lecture. In other words, if participants learn about each
lecture's content when they watch each lecture, the knowledge estimates should
capture that. After watching the \textit{Four Fundamental Forces} lecture,
participants should exhibit more knowledge for the content of that lecture than
they had before, and that knowledge should persist for the remainder of the
experiment. Specifically, knowledge about that lecture's content should be
relatively low when estimated using Quiz~1 responses, but should increase when
estimated using Quiz~2 or 3 responses (Fig.~\ref{fig:knowledge-timeseries}B).
Indeed, we found that participants' estimated knowledge about the content of
the \textit{Four Fundamental Forces} was substantially higher on Quiz~2 versus
Quiz~1 ($t(49) = 8.764,~p < 0.001$) and on Quiz~3 versus Quiz~1 ($t(49) =
10.519,~p < 0.001$). We found no reliable differences in estimated knowledge
about that lecture's content on Quiz~2 versus 3 ($t(49) = 0.160,~p = 0.874$).
Similarly, we hypothesized (and subsequently confirmed) that participants
should show greater estimated knowledge about the content of the \textit{Birth of
Stars} lecture after (versus before) watching it
(Fig.~\ref{fig:knowledge-timeseries}D). Specifically, since participants
watched that lecture after taking Quiz~2 (but before Quiz~3), we hypothesized
that their knowledge estimates should be relatively low on Quizzes~1 and 2, but
should show a ``boost'' on Quiz~3. Consistent with this prediction, we found no
reliable differences in estimated knowledge about the \textit{Birth of Stars}
lecture content on Quizzes~1 versus 2 ($t(49) = 1.013,~p = 0.316$), but the
estimated knowledge was substantially higher on Quiz~3 versus 2 ($t(49) =
10.561,~p < 0.001$) and Quiz~3 versus 1 ($t(49) = 8.969,~p < 0.001$).

Additionally, if we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should carry some
predictive information about whether the participant is likely to answer that
question correctly or incorrectly. To further validate our knowledge estimates, we developed a statistical approach to test this claim. 
For each question in turn, we used equation~\ref{eqn:prop} to estimate each 
participant's knowledge at the given question's embedding space coordinate, 
using all \textit{other} questions that participant answered on the same quiz.
%For each question in turn, for each
%participant, we used Equation~\ref{eqn:prop} to estimate (using all
%\textit{other} questions from the same quiz, from the same participant) the
%participant's knowledge at the held-out question's embedding coordinate. 
For
each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of \textit{correctly} answered
questions, and another for the estimated knowledge at the coordinates of
\textit{incorrectly} answered questions (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be applied to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with $k = 100$ topics. We hoped that increasing the number of
topics from 15 to 100 might help us to generalize the knowledge predictions.
(Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.)
As in our other analyses, we resampled each lecture's topic trajectory to 1~Hz
and projected each question into a shared text embedding space.
